---
title: Static plots - Assignments
output: html_document
---

### References and datasets

-   Original paper: Abbas et al. (2019) *Ecological Indicators* (<https://doi.org/10.1016/j.ecolind.2019.105524>)
-   StatQuest: PCA and other videos (<https://statquest.org/>)
-   Also check the lecture slides for L02-L07

# 1. Student ID and Name

## [Updated: You do not need to write your student ID and name]{style="color:red"}

# 2. Generative AI declaration

Students may use generative AI tools during assignment, for example, to check and improve code, provided that the content is original and reflects the student's own contributions. If generative AI is used, students must declare its use in the box below. An example statement is: “*During the assignment, I used [XXXX] to [XXXX]. I subsequently reviewed and edited the code and take(s) full responsibility for the final version of the code.”*

::: textbox
xxxxxxxxxxxxxx
:::

# 3. Assignments

## Github: Create a Github account

-   Create your Github account.
-   Install Git on your computer as well (the latest version = 2.53.0).
-   Not mandatory, but please read the advice about usernames in the lecture.

## Github: Create your first repository

-   Repository name must be “DASC3240_Assignment1.”
-   Make the visibility “public” so I can check it.
-   The repository must have a README and a License file.
-   Include a screenshot of your shell window showing the version of Git on your computer.

## Tree community data in Hong Kong

Go to Canvas and download `Abbas_2019_TableS4_ed1.csv` and `Abbas_2019_TableS4_ed2.csv`.

-   `Abbas_2019_TableS4_ed1.csv` includes the plot information (metadata).
-   `Abbas_2019_TableS4_ed2.csv` includes dominant tree species data in the 28 forest plots.

1.  Calculate the total abundance of the dominant tree species in each forest plot, and visualize the relationship between the total abundance and forest age. You should use `AGE_MEDIAN` column in the `Abbas_2019_TableS4_ed1.csv` file as a forest age index.

2.  Visualize the dominant tree species compositions using dimension reduction techniques. For visualization, you must use different colors for different forest age groups.

    2-1. Perform PCA to visualize the tree species composition.\
    2-2. Perform NMDS to visualize the tree species composition.\

# 3. Marking criteria and important notes

[**Please check the lecture PDF "L08_Assignment-I.pdf" provided on Canvas.**]{style="color:darkred"}

# 4. Your answers

### 1. Calculate the total abundance of the dominant tree species in each forest plot, and visualize the relationship between the total abundance and forest age.

-   You should use `AGE_MEDIAN` column in the `Abbas_2019_TableS4_ed1.csv` file as a forest age index.
-   You can calculate **the sum of each column** in `Abbas_2019_TableS4_ed2.csv` to calculate the total abundance of the dominant tree species in each plot.
-   You will combine the data from two dataframes, so please check if the column names of `Abbas_2019_TableS4_ed2.csv` and the row names (indices) of `Abbas_2019_TableS4_ed1.csv` match (by using R script, not by visual inspection!).
-   To visualize the relationship, barplot is **NOT** recommended.

```{r}
library(readr)
library(dplyr)
library(ggplot2)
# AI for revising code and grammar of comments

# ── 2. Read data ────────────────────────────────────────────────────────────
ed1 <- read_csv("Abbas_2019_TableS4_ed1.csv", show_col_types = FALSE)
ed2 <- read_csv("Abbas_2019_TableS4_ed2.csv", show_col_types = FALSE)

# ── 3. Verify plot identifiers match ────────────────────────────────────────
plots_ed1 <- ed1$PLOT
plots_ed2 <- colnames(ed2)[-1]  # exclude tree_species column

if (identical(sort(plots_ed1), sort(plots_ed2))) {
  message("All ", length(plots_ed1), " plot codes match perfectly between the two files.")
} else {
  stop("Plot codes do NOT match – check file structure!")
}

# ── 4. Calculate total abundance per plot ───────────────────────────────────
# Sum across all species (rows) for each plot (columns after the first)
total_abund <- colSums(ed2[, -1, drop = FALSE])

abund_df <- data.frame(
  PLOT            = names(total_abund),
  Total_Abundance = as.numeric(total_abund),
  stringsAsFactors = FALSE
)

# ── 5. Combine with AGE_MEDIAN (and AGE_CATEGORY for optional coloring) ─────
plot_data <- ed1 %>%
  select(PLOT, AGE_MEDIAN, AGE_CATEGORY) %>%
  left_join(abund_df, by = "PLOT") %>%
  arrange(AGE_MEDIAN)   # nicer for visualization

# Quick sanity check
summary(plot_data$Total_Abundance)
# Min ≈ 81, Median ≈ 263, Max ≈ 435

# ── 6. Visualization: Scatter + smooth trend (preferred over boxplot) ───────
ggplot(plot_data, aes(x = AGE_MEDIAN, y = Total_Abundance)) +
  
  # Points: semi-transparent, sized for visibility
  geom_point(size = 3.3, alpha = 0.85, color = "#1f77b4", shape = 16) +
  
  # Smooth trend line + confidence band (loess is robust for small n=28)
  geom_smooth(method = "loess", span = 0.75,   # slightly wider smooth
              color = "#d62728", linewidth = 1.4, se = TRUE, alpha = 0.18) +
  
  # Labels & title
  labs(
    title    = "Total Abundance of Dominant Tree Species vs. Median Forest Age",
    subtitle = "28 forest plots in Hong Kong – abundance = summed individuals of listed dominant species",
    x        = "Median forest age (years)",
    y        = "Total tree abundance (individuals)",
    caption  = "Data source: Abbas (2019) supplementary tables"
  ) +
  
  # Theme
  theme_minimal(base_size = 14) +
  theme(
    plot.title      = element_text(face = "bold", hjust = 0.5, size = 15),
    plot.subtitle   = element_text(hjust = 0.5, color = "grey40", size = 11),
    axis.title      = element_text(face = "bold"),
    panel.grid.major = element_line(color = "grey92"),
    panel.grid.minor = element_blank(),
    plot.caption    = element_text(hjust = 1, color = "grey50", size = 9)
  )
```

### 2. Visualize the dominant tree species compositions using PCA and NMDS.

There are a number of points you should consider:

-   For visualization, you must use different colors for different forest age groups. Use of different marker styles for different groups is also a good idea.
-   [**Set random seeds if your analysis includes random process(es)**]{style="color:darkred"}.
-   Before applying dimension reduction functions (e.g., `prcomp()`), please make sure that [**rows represent samples (= forest plots) and columns represent tree species**]{style="color:darkred"}. You have to transpose the data frame (In `Abbas_2019_TableS4_ed2.csv`, columns represent samples and rows represent tree species).
-   Usually we should consider whether we must perform the data preprocessing before applying dimension reduction techniques (e.g., standardization). [**In this assignment, please convert your data into compositional data using the following code**]{style="color:darkred"} `d_tree_rel <- apply(d_tree, 2, function(x) x/sum(x))` (`d_tree` represents the name of your data.frame. You should change it depending on your R object name). This means that, in this assignment, we are interested in tree community compositions, not abundances.

#### 2-1. Perform PCA to visualize the tree species composition.

After PCA, please visualize the result.

```{r}
# AI for revising code and grammar of comments
species_data <- t(ed2[, -1])
colnames(species_data) <- ed2$tree_species
rownames(species_data) <- plots_ed2

# Perform PCA with scaling (standardization)
pca <- prcomp(species_data, scale. = TRUE)

# Extract the first two principal components
pca_scores <- as.data.frame(pca$x[, 1:2])
pca_scores$PLOT <- rownames(pca_scores)

# Merge with age category from ed1
merged_pca <- merge(pca_scores, ed1[, c("PLOT", "AGE_CATEGORY")], by = "PLOT")

# Visualize PCA with different colors for age groups
ggplot(merged_pca, aes(x = PC1, y = PC2, color = AGE_CATEGORY)) +
  geom_point(size = 4) +  # Points representing each plot
  scale_color_manual(values = c("LT14" = "purple", "LT26" = "blue", "LT52" = "green", "LT70" = "orange", "GT70" = "red")) +  # Custom colors for age groups
  labs(title = "PCA of Dominant Tree Species Composition",
       x = "Principal Component 1",
       y = "Principal Component 2",
       color = "Age Category") +
  theme_minimal() +  # Clean theme
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),  # Center and bold title
        axis.title = element_text(face = "bold"),  # Bold axis labels
        legend.position = "right",  # Legend on the right
        legend.title = element_text(face = "bold"))  # Bold legend title

```

#### 2-2. Perform NMDS to visualize the tree species composition.

After NMDS, please visualize the result. You can use Bray-Curtis dissimilarity to calculate the dissimilarity among plots.

```{r}
# AI for revising code and grammar of comments
library(vegan)

# Use the species_data from above (plots as rows, species as columns)

# Set seed for reproducibility
set.seed(42)

# Perform NMDS using Bray-Curtis dissimilarity
nmds <- metaMDS(species_data, distance = "bray", k = 2, trymax = 100)  # k=2 for 2D, increase trymax if needed for convergence

# Extract NMDS scores
nmds_scores <- as.data.frame(scores(nmds, display = "sites"))
nmds_scores$PLOT <- rownames(nmds_scores)

# Merge with age category from ed1
merged_nmds <- merge(nmds_scores, ed1[, c("PLOT", "AGE_CATEGORY")], by = "PLOT")

# Visualize NMDS with different colors for age groups
ggplot(merged_nmds, aes(x = NMDS1, y = NMDS2, color = AGE_CATEGORY)) +
  geom_point(size = 4) +  # Points representing each plot
  scale_color_manual(values = c("LT14" = "purple", "LT26" = "blue", "LT52" = "green", "LT70" = "orange", "GT70" = "red")) +  # Custom colors for age groups
  labs(title = "NMDS of Dominant Tree Species Composition",
       x = "NMDS Axis 1",
       y = "NMDS Axis 2",
       color = "Age Category") +
  theme_minimal() +  # Clean theme
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),  # Center and bold title
        axis.title = element_text(face = "bold"),  # Bold axis labels
        legend.position = "right",  # Legend on the right
        legend.title = element_text(face = "bold"))  # Bold legend title

```
